Dynamics of thin liquid film flows studied with the lattice Boltzmann method

### 2023-08-10 chapters/Paper_five.tex
All references polished, conclusion missing

More generally, droplet coalescence has attracted much attention in the past two decades, see refs.~\cite{eggers1999coalescence, duchemin_eggers_josserand_2003, PhysRevLett.95.164503, PhysRevLett.106.114501, doi:10.1063/1.4828721}. 
The driving forces during coalescence are, the minimization of surface area and the minimization of interfacial curvature. 
These arguments hold true for scenarios beyond simple Newtonian liquids, in fact they are applicable to liquid lens coalescence~\cite{PhysRevLett.124.194502} or coalescence of quasi 2D liquids~\cite{klopp2020self, doi:10.1021/acs.langmuir.0c02139}.
What is common to the above-mentioned examples is the equilibrium state, which is one where the two droplets have coalesced.
This is of course in agreement with a minimized surface area and minimal curvature.

However, this reasoning doesn't always apply. 
Recently, Kern et al. have shown that viscoplastic effects can arrest the coalescence, yielding a stable twin drop states~\cite{PhysRevFluids.7.L081601}.
What has been known for some time, however, is that a surface tension gradient influences drop coalescence, as demonstrated by Riegler and Lazar and later Karpitschka et al.~\cite{PhysRevLett.109.066103, doi:10.1021/la500459v, karpitschka2014sharp, bruning2018delayed}.
They performed experiments with two drops from different but miscible liquids, therefore having a sharp surface tension gradient from one drop to the other.
Bocia and Bestehorn showed non-coalescence of droplets with numerical simulations~\cite{PhysRevE.82.036312, borcia2011coalescence}.
The clean formulation and derivation was later done by Karpitschka et al.~\cite{PhysRevLett.109.066103}, as they identify an effective Marangoni flow that stabilizes the two droplet system.
We revisit that problem with numerical simulations using a lattice Boltzmann method that is based on the thin film equation~\cite{PhysRevE.100.033313}.
First, we move to the regime where the droplets become tiny and the disjoining pressure can no longer be neglected.
Second, we address the impact of the surface tension beyond the reduction to its absolute contrast.
We ask the straightforward question, if the concrete function of $\gamma(x)$ can turn non-coalescing states into coalescing ones.
Within this framework, we discuss the growth law of the liquid bridge and the influence of the disjoining pressure on coalescence.

@@

\section{Theory}\label{sec:theory_five}
The theoretical approach we use to study this system is the lubrication approximation~\cite{Reynolds, oronLongscaleEvolutionThin1997, diezGlobalModelsMoving2000}.
Applying this approximation to the Navier-Stokes equation yields the thin film equation, which for a singular horizontal dimension reads~\cite{bonnWettingSpreading2009, crasterDynamicsStabilityThin2009, thielePatternedDepositionMoving2014}
\begin{equation}\label{eq:thin_film_simple}
    \partial_t h = \partial_x \left(\frac{h^3}{3\mu}\partial_x p\right),
\end{equation}
where $\frac{h^3}{3\mu}$ is the mobility $M(h)$\footnote{$m(h) = h^3/3\mu$ is true only for a no-slip velocity boundary condition.}, $h(x,t)$ is the thickness of the film at time $t$ and position $x$, $\mu$ is the liquid's viscosity, and $p$ it's pressure.
The pressure accounts for both the surface tension and the correct fluid substrate behavior (wetting)~\cite{PhysRevE.100.033313}.
We therefore write the pressure as
\begin{equation}\label{eq:pressure_five}
    p = \gamma \partial_x^2 h + \Pi(h),
\end{equation}
where the first term is the (1D) Laplacian of the liquid-vapor interface and $\Pi(h)$ is the disjoining pressure~\cite{oronLongscaleEvolutionThin1997, bonnWettingSpreading2009, peschkaSignaturesSlipDewetting2019, diezGlobalModelsMoving2000},
\begin{equation}\label{eq:disjoin}
    \Pi(h) = K(\gamma,\theta)\left[\left(\frac{h_{\ast}}{h}\right)^n - \left(\frac{h_{\ast}}{h}\right)^m\right].
\end{equation}
The prefactor $K(\gamma,\theta)\propto \gamma(1-\cos(\theta))$ encodes the wettability and as such the equilibrium contact angle $\theta_{\text{eq.}}$.
This prefactor can be understood as a linking factor to the Hamaker constant ($\mathcal{A}$) with~~\cite{diezMetallicthinfilmInstabilitySpatially2016, bestehorn20033d, van1988interfacial}
\begin{equation}
    \mathcal{A} = 6\pi h_{\ast}^3 K(\theta).    
\end{equation}
Additionally, $h_{\ast}$ defines the thickness of the precursor layer, therefore the thickness where $\Pi(h_{\ast}) = 0$.
The pair of powers $(n,m)$ need to satisfy $n > m$ and $m > 1$.
For an in deep presentation of the derivation we refer to the work by Schwartz and Eley and Bonn et al.~\cite{schwartzSimulationDropletMotion1998, bonnWettingSpreading2009}.

@@

\subsection{Flows}\label{subsec:flows_theory}
Eq.~(\ref{eq:thin_with_marangoni}) defines the dynamics of the system we are interested in.
Before dealing with the full complexity, we take a step back and set $\partial_x\gamma = 0$.
In this case, we know that the two touching droplets will coalesce.
Following the argumentation of Riegler, Lazar and Eddie et al.~\cite{doi:10.1021/la800630w, PhysRevLett.111.144502},
\begin{equation}\label{eq:cap_pressure}
    P_{\text{cap.}} \sim\gamma\kappa \sim \frac{\gamma}{h_0},
\end{equation}
where $P_{\text{cap.}}$ is the capillary pressure and $\kappa$ is the curvature of the liquid vapor interface.
Upon rescaling, Eddi et al. found in their analytical analysis that $h_0$ is the only relevant scale, justifying therefore $\kappa \sim 1/h_0$~\cite{PhysRevLett.111.144502}. 
The resulting flow towards the bridge induces an inertial or ``Bernoulli'' pressure
\begin{equation}\label{eq:P_Bernulli}
    P_{\text{iner.}} \sim \rho\left(\frac{h_0}{t}\right)^2.
\end{equation}
Balancing Eqs.~(\ref{eq:cap_pressure}-\ref{eq:P_Bernulli}) and solving for $h_0$ yields a growth law 
\begin{equation}\label{eq:coal_powerlaw}
    h_0(t) \sim t^{2/3},
\end{equation}
for contact angles below $\theta < \pi/2$~\cite{PhysRevLett.111.144502, keller2002breaking}.
Recently, a similar growth law for the coalescence of low viscosity liquid lenses has been experimentally observed and theoretically validated, see Hack et al.~\cite{PhysRevLett.124.194502}.

@@

\section{Method}\label{sec:method_five}
\begin{figure}
    \centering
    \includegraphics[width=0.68\textwidth]{graphics/gammas.pdf}
    \caption{Surface tension fields $\gamma(x)$ normalized with $\gamma_0$.
    Different colors and symbols display different surface tensions with $x=0$ is the coordinate of the bridge at $t=0$.
    The surface tension functions are given by Eqs.~(\ref{eq:gamma_const}-\ref{eq:gamma_tanh}).
    }
    \label{fig:gammas}
\end{figure}
We use a computational method, namely the lattice Boltzmann method (LBM), to iteratively solves Eq.~(\ref{eq:thin_with_marangoni}).
The details of the method can be found in the App.~\ref{app:two} and in the references~\cite{PhysRevE.100.033313, zitzLatticeBoltzmannSimulations2021}.

@@

Performing the limits discussed in ref.~\cite{PhysRevE.100.033313, zitzLatticeBoltzmannSimulations2021}, this system becomes an effective solver for Eq~(\ref{eq:thin_with_marangoni}).
Due to the inclusion of $\Pi(h)$, this method allows us to test the non-coalescence criteria for smaller scales, well below $1mm$ droplet diameter~\cite{doi:10.1021/la500459v, karpitschka2014sharp}.

@@

\section{Results}\label{sec:results_five}
\begin{figure*}
    \centering
    \includegraphics[width=1.0\textwidth]{graphics/height_three_times_three_gamma.pdf}
    \caption{Snapshots of the thickness $h(x)$ taken at three different time steps.
    The three time step have different marks, namely bullets, stars and triangles for $t=0.14\tau$, $t=14\tau$ and $t=140\tau$ respectively.
    In panel (a) we show the evolution for a constant surface tension, see Fig.~\ref{fig:gammas} blue bullets.
    In (b) and (c) we use Eq.~(\ref{eq:gamma_tanh}) with $w =5$ and $w=100$ respectively. 
    }
    \label{fig:final_state}
\end{figure*}
All numerical experiments start with the same initial condition, two barely overlapping circular segments (equivalent to spherical caps in three dimensions) as illustrated in Fig.~\ref{fig:schematics}.
We use the radius of the initial droplets $R_0$ to normalize length scales~\cite{PhysRevLett.111.144502, PhysRevLett.95.164503}.
As a characteristic time scale we use the inertio-capillary time, 
\begin{equation}\label{eq:inertio-cap-time}
    \tau = \sqrt{\frac{\rho R_0^3}{\gamma}}.
\end{equation}

@@

This behavior is more pronounced for larger smearing values $w$.
Increasing $w$ yields smaller $\partial_x\gamma$ values at the bridge, as can be computed using Eq.~(\ref{eq:gamma_tanh}). 
For $w\gtrsim R_0$ we no longer see a separation, but an asymmetric coalescence.
A simpler measure to assess this statement is the difference in droplet height,
\begin{equation}\label{eq:drop_diff_h}
    \Delta h = \left|h_{d,1} - h_{d,2}\right|
\end{equation}
where $h_{d,i}$ with $i=1,2$ identifies the left and right droplet, respectively.
These measurements are shown in Fig.~\ref{fig:drop_diff} for a subset of chosen surface tension fields.
For $w < 50$ (or $w/R_0 \le 1/3$) we observe a collapse of the data on short time scales.
The same curves, leaving aside the constant surface tension, show a kink in their slope at what we call the separation time $\tau_s$, which is defined according to $\min_t(h_0(\tau_{s}) \le h_{\ast})$ and shown in the inset of Fig~\ref{fig:drop_diff}.
On top of our data, the blue bullets, we plot a power-law $t \sim w^{3/2}$.
The motivation for this relation is as follows
\begin{equation}\label{eq:velsim1}
    \frac{\gamma}{\mu} \sim \frac{h_0}{2\mu}\partial_x\gamma,
\end{equation}
where $\gamma/\mu$ is a fluid specific capillary velocity and the left hand side is due to Marangoni~\cite{PhysRevLett.95.164503, doi:10.1021/la971292t}.
Rearranging the above similarity to have the bridge height on the left yields
\begin{equation}\label{velsim2}
    h_0 \sim \frac{2w\gamma}{\Delta\gamma},
\end{equation}
where we used $\partial_x\gamma \sim \Delta\gamma/w$, see Eq.~(\ref{eq:gamma_tanh}).
From Fig.~\ref{fig:bridge_growth} we know that Eq.~(\ref{eq:coal_powerlaw}) holds true for most of our data, therefore
\begin{equation}\label{velsim3}
    \tau_s \sim \sqrt{\frac{8\gamma^3}{(\Delta\gamma)^3}w^3}.
\end{equation}
The black dashed line in the inset of Fig.~\ref{fig:drop_diff} is given by $\sim w^{3/2}$, showing good agreement with the data.

>>>>

More generally, droplet coalescence has attracted much attention in the past two decades, see refs.~\cite{eggersCoalescenceLiquidDrops1999, duchemin_eggers_josserand_2003, aartsHydrodynamicsDropletCoalescence2005, PhysRevLett.106.114501, doi:10.1063/1.4828721}. 
The driving forces during coalescence are, the minimization of surface area and the minimization of interfacial curvature. 
These arguments hold true for scenarios beyond simple Newtonian liquids, in fact they are applicable to liquid lens coalescence~\cite{PhysRevLett.124.194502} or coalescence of quasi 2D liquids~\cite{klopp2020self, doi:10.1021/acs.langmuir.0c02139}.
What is common to the above-mentioned examples is the equilibrium state, which is one where the two droplets have coalesced.
This is of course in agreement with a minimized surface area and minimal curvature.

However, this reasoning doesn't always apply. 
Recently, Kern et al. have shown that viscoplastic effects can arrest the coalescence, yielding a stable twin drop states~\cite{PhysRevFluids.7.L081601}.
What has been known for some time, however, is that a surface tension gradient influences drop coalescence, as demonstrated by Riegler and Lazar and later Karpitschka et al.~\cite{PhysRevLett.109.066103, doi:10.1021/la500459v, karpitschka2014sharp, bruning2018delayed}.
They performed experiments with two drops from different but miscible liquids, therefore having a sharp surface tension gradient from one drop to the other.
Bocia and Bestehorn showed non-coalescence of droplets with numerical simulations~\cite{PhysRevE.82.036312, borcia2011coalescence}.
The clean formulation and derivation was later done by Karpitschka et al.~\cite{PhysRevLett.109.066103}, as they identify an effective Marangoni flow that stabilizes the two droplet system.
We revisit that problem with numerical simulations using a lattice Boltzmann method that is based on the thin film equation~\cite{zitzLatticeBoltzmannMethod2019}.
First, we move to the regime where the droplets become tiny and the disjoining pressure can no longer be neglected.
Second, we address the impact of the surface tension beyond the reduction to its absolute contrast.
We ask the straightforward question, if the concrete function of $\gamma(x)$ can turn non-coalescing states into coalescing ones.
Within this framework, we discuss the growth law of the liquid bridge and the influence of the disjoining pressure on coalescence.

@@

\section{Theory}\label{sec:theory_five}
The theoretical approach we use to study this system is the lubrication approximation~\cite{Reynolds, oronLongscaleEvolutionThin1997, diezGlobalModelsMoving2000}.
Applying this approximation to the Navier-Stokes equation yields the thin film equation, which for a singular horizontal dimension reads~\cite{bonnWettingSpreading2009, crasterDynamicsStabilityThin2009, thielePatternedDepositionMoving2014}
\begin{equation}\label{eq:thin_film_simple}
    \partial_t h = \partial_x \left(\frac{h^3}{3\mu}\partial_x p\right),
\end{equation}
where $\frac{h^3}{3\mu}$ is the mobility $M(h)$\footnote{$m(h) = h^3/3\mu$ is true only for a no-slip velocity boundary condition.}, $h(x,t)$ is the thickness of the film at time $t$ and position $x$, $\mu$ is the liquid's viscosity, and $p$ it's pressure.
The pressure accounts for both the surface tension and the correct fluid substrate behavior (wetting)~\cite{zitzLatticeBoltzmannMethod2019}.
We therefore write the pressure as
\begin{equation}\label{eq:pressure_five}
    p = \gamma \partial_x^2 h + \Pi(h),
\end{equation}
where the first term is the (1D) Laplacian of the liquid-vapor interface and $\Pi(h)$ is the disjoining pressure~\cite{oronLongscaleEvolutionThin1997, bonnWettingSpreading2009, peschkaSignaturesSlipDewetting2019, diezGlobalModelsMoving2000},
\begin{equation}\label{eq:disjoin}
    \Pi(h) = K(\gamma,\theta)\left[\left(\frac{h_{\ast}}{h}\right)^n - \left(\frac{h_{\ast}}{h}\right)^m\right].
\end{equation}
The prefactor $K(\gamma,\theta)\propto \gamma(1-\cos(\theta))$ encodes the wettability and as such the equilibrium contact angle $\theta_{\text{eq.}}$.
This prefactor can be understood as a linking factor to the Hamaker constant ($\mathcal{A}$) with~~\cite{diezMetallicthinfilmInstabilitySpatially2016, bestehorn20033d, van1988interfacial}
\begin{equation}
    \mathcal{A} = 6\pi h_{\ast}^3 K(\theta).    
\end{equation}
Additionally, $h_{\ast}$ defines the thickness of the precursor layer, therefore the thickness where $\Pi(h_{\ast}) = 0$.
The pair of powers $(n,m)$ need to satisfy $n > m$ and $m > 1$.
For an in deep presentation of the derivation we refer to the work by Schwartz and Eley and Bonn et al.~\cite{schwartzSimulationDropletMotion1998, bonnWettingSpreading2009}.

@@

\subsection{Flows}\label{subsec:flows_theory}
Eq.~(\ref{eq:thin_with_marangoni}) defines the dynamics of the system we are interested in.
Before dealing with the full complexity, we take a step back and set $\partial_x\gamma = 0$.
In this case, we know that the two touching droplets will coalesce.
Following the argumentation of Riegler, Lazar and Eddie et al.~\cite{doi:10.1021/la800630w, eddiInfluenceDropletGeometry2013},
\begin{equation}\label{eq:cap_pressure}
    P_{\text{cap.}} \sim\gamma\kappa \sim \frac{\gamma}{h_0},
\end{equation}
where $P_{\text{cap.}}$ is the capillary pressure and $\kappa$ is the curvature of the liquid vapor interface.
Upon rescaling, Eddi et al. found in their analytical analysis that $h_0$ is the only relevant scale, justifying therefore $\kappa \sim 1/h_0$~\cite{eddiInfluenceDropletGeometry2013}. 
The resulting flow towards the bridge induces an inertial or ``Bernoulli'' pressure
\begin{equation}\label{eq:P_Bernulli}
    P_{\text{iner.}} \sim \rho\left(\frac{h_0}{t}\right)^2.
\end{equation}
Balancing Eqs.~(\ref{eq:cap_pressure}-\ref{eq:P_Bernulli}) and solving for $h_0$ yields a growth law 
\begin{equation}\label{eq:coal_powerlaw}
    h_0(t) \sim t^{2/3},
\end{equation}
for contact angles below $\theta < \pi/2$~\cite{eddiInfluenceDropletGeometry2013, keller2002breaking}.
Recently, a similar growth law for the coalescence of low viscosity liquid lenses has been experimentally observed and theoretically validated, see Hack et al.~\cite{PhysRevLett.124.194502}.

@@

\section{Method}\label{sec:method_five}
\begin{figure}
    \centering
    \includegraphics[width=0.68\textwidth]{graphics/gammas.pdf}
    \caption{Surface tension fields $\gamma(x)$ normalized with $\gamma_0$.
    Different colors and symbols display different surface tensions with $x=0$ is the coordinate of the bridge at $t=0$.
    The surface tension functions are given by Eqs.~(\ref{eq:gamma_const}-\ref{eq:gamma_tanh}).
    }
    \label{fig:gammas}
\end{figure}
We use a computational method, namely the lattice Boltzmann method (LBM), to iteratively solves Eq.~(\ref{eq:thin_with_marangoni}).
The details of the method can be found in the App.~\ref{app:two} and in the references~\cite{zitzLatticeBoltzmannMethod2019, zitzLatticeBoltzmannSimulations2021}.

@@

Performing the limits discussed in ref.~\cite{zitzLatticeBoltzmannMethod2019, zitzLatticeBoltzmannSimulations2021}, this system becomes an effective solver for Eq~(\ref{eq:thin_with_marangoni}).
Due to the inclusion of $\Pi(h)$, this method allows us to test the non-coalescence criteria for smaller scales, well below $1mm$ droplet diameter~\cite{doi:10.1021/la500459v, karpitschka2014sharp}.

@@

\section{Results}\label{sec:results_five}
\begin{figure*}
    \centering
    \includegraphics[width=1.0\textwidth]{graphics/height_three_times_three_gamma.pdf}
    \caption{Snapshots of the thickness $h(x)$ taken at three different time steps.
    The three time step have different marks, namely bullets, stars and triangles for $t=0.14\tau$, $t=14\tau$ and $t=140\tau$ respectively.
    In panel (a) we show the evolution for a constant surface tension, see Fig.~\ref{fig:gammas} blue bullets.
    In (b) and (c) we use Eq.~(\ref{eq:gamma_tanh}) with $w =5$ and $w=100$ respectively. 
    }
    \label{fig:final_state}
\end{figure*}
All numerical experiments start with the same initial condition, two barely overlapping circular segments (equivalent to spherical caps in three dimensions) as illustrated in Fig.~\ref{fig:schematics}.
We use the radius of the initial droplets $R_0$ to normalize length scales~\cite{eddiInfluenceDropletGeometry2013, aartsHydrodynamicsDropletCoalescence2005}.
As a characteristic time scale we use the inertio-capillary time, 
\begin{equation}\label{eq:inertio-cap-time}
    \tau = \sqrt{\frac{\rho R_0^3}{\gamma}}.
\end{equation}

@@

This behavior is more pronounced for larger smearing values $w$.
Increasing $w$ yields smaller $\partial_x\gamma$ values at the bridge, as can be computed using Eq.~(\ref{eq:gamma_tanh}). 
For $w\gtrsim R_0$ we no longer see a separation, but an asymmetric coalescence.
A simpler measure to assess this statement is the difference in droplet height,
\begin{equation}\label{eq:drop_diff_h}
    \Delta h = \left|h_{d,1} - h_{d,2}\right|
\end{equation}
where $h_{d,i}$ with $i=1,2$ identifies the left and right droplet, respectively.
These measurements are shown in Fig.~\ref{fig:drop_diff} for a subset of chosen surface tension fields.
For $w < 50$ (or $w/R_0 \le 1/3$) we observe a collapse of the data on short time scales.
The same curves, leaving aside the constant surface tension, show a kink in their slope at what we call the separation time $\tau_s$, which is defined according to $\min_t(h_0(\tau_{s}) \le h_{\ast})$ and shown in the inset of Fig~\ref{fig:drop_diff}.
On top of our data, the blue bullets, we plot a power-law $t \sim w^{3/2}$.
The motivation for this relation is as follows
\begin{equation}\label{eq:velsim1}
    \frac{\gamma}{\mu} \sim \frac{h_0}{2\mu}\partial_x\gamma,
\end{equation}
where $\gamma/\mu$ is a fluid specific capillary velocity and the left hand side is due to Marangoni~\cite{aartsHydrodynamicsDropletCoalescence2005, doi:10.1021/la971292t}.
Rearranging the above similarity to have the bridge height on the left yields
\begin{equation}\label{velsim2}
    h_0 \sim \frac{2w\gamma}{\Delta\gamma},
\end{equation}
where we used $\partial_x\gamma \sim \Delta\gamma/w$, see Eq.~(\ref{eq:gamma_tanh}).
From Fig.~\ref{fig:bridge_growth} we know that Eq.~(\ref{eq:coal_powerlaw}) holds true for most of our data, therefore
\begin{equation}\label{velsim3}
    \tau_s \sim \sqrt{\frac{8\gamma^3}{(\Delta\gamma)^3}w^3}.
\end{equation}
The black dashed line in the inset of Fig.~\ref{fig:drop_diff} is given by $\sim w^{3/2}$, showing good agreement with the data.
